---
title: GDP vs Life Exp - CO vs AU comparision last 70 years
author: Erika Muñoz
date: 2026-07-16
echo: false
warning: false
format: html
theme: lux
categories: [R, 26Winter, "data: gapminder.csv"]
---
## **Dataset**
Selected dataset is gapminder but with 2015 to 2025 addition of Data from the World Bank Open Data 

```{r}
gapminder <- read.csv("./gapminder.csv")
```

After that I make sure to charge all the libraries I used to processing the data 

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(skimr)
```

Then, the data was filtered to the selected countries for comparision and filtered for the relevant variables. 

```{r}
AUvsCO <- filter(gapminder,country%in% c("Australia", "Colombia"))

AUvsCO_small <- select(AUvsCO, -continent,-pop) 
write.csv(AUvsCO_small, "./AUvCO.csv")
```

After that I added the new data to the set. 

```{r}

new_row <- data.frame(
country = "Australia",
year = 2018,
lifeExp = 83.0,
gdpPercap = 57410)
AUvsCO_small <- rbind(AUvsCO_small, new_row)

nuevos_datos <- data.frame(
  country = c(
    "Australia", "Australia", "Australia", "Australia",
    "Colombia", "Colombia", "Colombia", "Colombia", "Colombia"
  ),
  year = c(
    2019, 2020, 2021, 2025,
    2018, 2019, 2020, 2021, 2025
  ),
  lifeExp = c(
    83, 83, 83.3, 84.18,
    76.6,  76.8, 74.8, 72.7, 78.1
  ),
  gdpPercap = c(
    55194, 51983, 64335, 66350,
    6817, 6473, 5340, 5280, 8562
  ))

AUvsCO_small <- rbind(AUvsCO_small, nuevos_datos)

tail(AUvsCO_small)

```
## **Context**
Both countries have different socioeconomical context which implies serious differences in the data. I am focusing in a long term comparision plus focusing in something that affected all the world, being COVID-19 Pandemic. 

## **Visualisation**

Correlation between life expectancy and GDP per capita 

```{r}
ggplot(AUvsCO_small,
       aes(x = gdpPercap,
           y = lifeExp,
           color = country)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  labs(
    title = "GDP per Capita and Life Expectancy",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy"
  ) +
  theme_minimal()
```

I also transform some data to percentage to make easier comparisions through years. 

```{r}
AU_growth <- AUvsCO_small %>%
  group_by(country) %>%
  arrange(year) %>%
  mutate(
    GDP_index = gdpPercap / first(gdpPercap) * 100
  )

Life_Change <- AUvsCO_small %>%
  group_by(country) %>%
  arrange(year) %>%
  mutate(
    LE_index = lifeExp / first(lifeExp) * 100
  )


```


Life expectancy comparision 

```{r}
ggplotly(ggplot(AUvsCO_small,
       aes(year, lifeExp, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#e280d0", "#96e3e1")) +
  labs(
    title = "Life Expectancy",
    subtitle = "1952-2025",
    x = NULL,
    y = "Years",
    color = "Country"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "top"
  ))
```

GDP per Capita comparision

```{r}
ggplotly(ggplot(AUvsCO_small,
       aes(year, gdpPercap, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#e280d0", "#96e3e1")) +
  labs(
    title = "GDP",
    subtitle = "1952-2025",
    x = NULL,
    y = "gdp",
    color = "Country"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "top"
  ))

```

## **Stats**
Moderate to strong correlation between economic status or capability of a country and the life expectancy of its population. 

```{r}
model <- lm(lifeExp ~ gdpPercap + country,
            data = AUvsCO_small)

summary(model)
summary(model)$r.squared

AUvsCO_small %>%
  group_by(country) %>%
  summarise(
    correlation = cor(gdpPercap, lifeExp)
  )

ggplot(AUvsCO_small,
aes(country, lifeExp,
fill = country)) +
geom_boxplot()

ggplot(AUvsCO_small,
aes(country, gdpPercap,
fill = country)) +
geom_boxplot()

```

