---
title: "FC Barcelona Team Analysis"
author: "Adriel Wijaya"
format: html
editor: visual
date: today
#bibliography: references.bib
---

## Background

**FC Barcelona** (Futbol Club Barcelona ) is a football club founded in 1899 by a group of Swiss, Catalan, German, and English footballers led by Joan Gamper[@fcbarce2025]. The club motto "Més que un club" meaning "More than a club"[@fcbarce2025].

## FC Barcelona Player 2023-2024

Table below shows football player that plays for FC Barcelona.

```{r}
library(dplyr)
#|eval: False
```

```{r, fig.cap="Table.1 FC Barcelona Player Season 2023-2024", fig.align='center'}

dataset <- read.csv("../../data_sources/Players2024.csv")
subset <- dataset |> filter(club=="Futbol Club Barcelona")
subset <- subset |> select(-club)

colnames(subset) <- c("Name", "Birth Date", "Height", "Positions","Nationality","Age")#change the column name

subset |> knitr::kable("html")
```

## La Masia

FC Barcelona's success is contributed by it's youth football academy called **La Masia** (The Farmhouse). It is considered as the best youth system in world football. It can be seen in the figure below that the majority of FC Barcelona players from Spain.

```{r, fig.cap="Figure 1. FC Barcelona Player Based on Nationality and Positions", fig.align='center'}
library(ggplot2)
subset |> ggplot(mapping=aes(x=Positions,fill=Nationality))+
  geom_bar()
```

This is further shown in the figure below where for every position (except Goalkeeper) FC Barcelona has a young player (age below 25) coming from La Masia.

```{r, fig.cap="Figure 2. FC Barcelona Player Based on Age, Nationality and Positions", fig.align='center'}
subset |> ggplot(aes(x=Age, 
                     fill=Nationality))+
  geom_histogram(binwidth = 1.5)+
  scale_fill_manual(values=c("#a50044","#004d98","#edbb00","#ffed02","#db0030","#157b6d","#456789"))+
  facet_wrap(vars(Positions))
```

The success of FC Barcelona in the footballing world shows the importance of developing young talent through club academy. This not only beneficial for the club but also the player.