---
title: "Taste testing coffee results"
author: "zoë"
format: html
editor: visual


---
## <span style="color:#6F4E37;"> Does your age impact the way you like to make your coffee? </span>


<span style="color:#8B5A2B;"> 
This plot shows age group, preference of coffee making method, and the number of cups of coffee an individual generally consumes.</span>

```{r}
#| echo: false
#| include: false

 library(dplyr)
 library(readr)
 library(ggplot2)
 library(plotly)
 library(tidyr)
```


```{r}
#| echo: false
#| message: false

#### Project 1 - bringing in coffee data #####

coffee <- read_csv("data/coffee_survey.csv")

coffee_age_plot_1 <- ggplot(
  coffee %>%
    filter(
      !is.na(cups),
      cups != "Less than 1"
    ),
  aes(x = age, y = favourite, color = cups)
) +
  geom_count() +
  scale_color_manual(
    name = "Cups per day",
    values = c(
      "1"           = "#CEBCB6",# light latte
      "2"           = "#AD8F85",
      "3"           = "#87655A",
      "4"           = "#493731",
      "More than 4" = "#0C0908" # dark espresso
    )
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplotly(coffee_age_plot_1)



```


## <span style="color:#6F4E37;"> What age group drinks the most cups of coffee? </span>

```{r}
#| echo: false
ggplot(coffee, aes(x = age, y = cups)) +
  geom_count()
```

<span style="color:#8B5A2B;"> Seems like 24-35 year olds usually have one or two cups of coffee (or, this means that there were more 24-35 year olds sampled than other age groups. Whoops.).</span>

## <span style="color:#6F4E37;"> Which of the blind-tasted coffees are percieved as the most acidic? </span>

```{r}
#| echo: false
##pivot

coffee_acidity_long <- coffee %>%
  pivot_longer(
    cols = starts_with("coffee_") & ends_with("_acidity"),
    names_to = "coffee",
    values_to = "acidity"
  )

# add labels

coffee_acidity_long <- coffee_acidity_long %>%
  mutate(
    coffee = recode(coffee,
                    coffee_a_acidity = "Coffee A",
                    coffee_b_acidity = "Coffee B",
                    coffee_c_acidity = "Coffee C",
                    coffee_d_acidity = "Coffee D"
    )
  )

# remove NA

coffee_acidity_long <- coffee_acidity_long %>%
  filter(!is.na(acidity))

# plot

ggplot(coffee_acidity_long, aes(x = coffee, y = acidity)) +
  geom_boxplot() +
  labs(
    x = "Coffee",
    y = "Acidity rating",
    title = "Perceived acidity of coffees A–D"
  ) +
  theme_minimal()
```

<span style="color:#8B5A2B;">Looks like coffee A and D were the most acidic, as voted by the taste-testers.</span>


## <span style="color:#6F4E37;"> Comparing acidity and bitterness, and showing popularity </span>


```{r}
#| echo: false
library(ggplot2)
library(dplyr)
library(plotly)


coffee_acidity_long <- coffee %>%
  pivot_longer(
    cols = starts_with("coffee_") & ends_with("_acidity"),
    names_to = "coffee",
    values_to = "acidity"
  )

# add labels

coffee_acidity_long <- coffee_acidity_long %>%
  mutate(
    coffee = recode(coffee,
                    coffee_a_acidity = "Coffee A",
                    coffee_b_acidity = "Coffee B",
                    coffee_c_acidity = "Coffee C",
                    coffee_d_acidity = "Coffee D"
    )
  )


# remove NA

coffee_acidity_long <- coffee_acidity_long %>%
  filter(!is.na(acidity))


# ## taking the mean of acidity
# 
# coffee_acidity_long %>%
#   group_by(coffee) %>%
#   summarise(
#     mean_acidity = mean(acidity),
#     se = sd(acidity) / sqrt(n()),
#     .groups = "drop"
#   ) %>%
#   ggplot(aes(coffee, mean_acidity)) +
#   geom_col() +
#   geom_errorbar(
#     aes(ymin = mean_acidity - se, ymax = mean_acidity + se),
#     width = 0.2
#   ) +
#   labs(y = "Mean acidity rating") +
#   theme_minimal()


coffee_scores <- coffee %>%
  pivot_longer(
    cols = matches("^coffee_[abcd]_(bitterness|acidity)$"),
    names_to = c("coffee", "metric"),
    names_pattern = "^coffee_([abcd])_(bitterness|acidity)$",
    values_to = "score"
  ) %>%
  filter(!is.na(score)) %>%
  mutate(coffee = paste("Coffee", toupper(coffee))) %>%
  group_by(coffee, metric) %>%
  summarise(
    mean_score = mean(score),
    se = sd(score) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = c(mean_score, se, n)
  )
coffee_scores <- coffee_scores %>%
  rename(
    mean_acidity    = mean_score_acidity,
    mean_bitterness = mean_score_bitterness
  )


## determine coffee popularity

coffee_pref_counts <- coffee %>%
  filter(!is.na(prefer_overall)) %>%
  count(prefer_overall, name = "n_prefer") %>%
  rename(coffee = prefer_overall)

# join to new coffee dataframe

coffee_scores <- coffee_scores %>%
  left_join(coffee_pref_counts, by = "coffee")




# label coffees with color instead of text

coffee_acid_bitter_pop <- ggplot(
  coffee_scores,
  aes(
    x = mean_acidity,
    y = mean_bitterness,
    size = n_prefer,
    color = coffee
  )
) +
  geom_point(alpha = 0.8)+
  scale_size(
    name = "Number of respondents\npreferring this coffee",
    range = c(4, 14)
  ) +
  scale_color_manual(
    name = "Coffee",
    values = c(
      "Coffee A" = "#6F4E37",  # classic coffee brown
      "Coffee B" = "#8B5A2B",  # lighter roast
      "Coffee C" = "#4B2E2A",  # dark roast
      "Coffee D" = "#A67C52"   # latte brown
    )
  ) +
  labs(
    x = "Mean perceived acidity",
    y = "Mean perceived bitterness",
    title = "Coffee flavor space (size = popularity)"
  ) +
  theme_minimal()

ggplotly(coffee_acid_bitter_pop)

```
