---
title: Fuel Price Analysis in QLD
author: Surya Chandra Rejinthala
date: 2026-07-16

---
## Introduction
This project analyzes Queensland fuel price data to identify pricing trends across fuel types and time The analysis was conducted using R programming.

## Importing Data and necessary libraries
```{r} 
#Importing data
fueldata <- read.csv("data/qld_fuel.csv")

#Loading packages
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(ggpubr)
```
## Overview of Fuel Types and total number of transcations

```{r}
fueldata <- fueldata |> 
select(-X, -X_id) #removing X and X_id column

#converting date from str to date
date <- fueldata$TransactionDateutc
date <- ymd_hms(date)
#unique Fueltypes or types of fuels
unique(fueldata$Fuel_Type)

ggplot(fueldata,
       aes(y= Fuel_Type))+
geom_bar(fill = "blue")+theme_minimal()
```
## Top 5 Fuel Brands

Available Fuel Brands in QLD and the Top 5 fuel brands with the highest number of transcations over the year are.
```{r}
#fuel Companies

unique(fueldata$Site_Brand)

#Top 5 brands 
top5 <- fueldata |> 
count(Site_Brand, sort= TRUE) |> 
 head(n = 5)

print(top5)

ggplot(top5,aes(x=Site_Brand, y=n))+geom_col(fill= "light green")+theme_minimal()
```

## Summary of Fuel Price
```{r}

summary(fueldata$Price)
mean(fueldata$Price)
median(fueldata$Price)
sd(fueldata$Price)
#Histogram of data
ggplot(fueldata,
aes(Price))+
geom_histogram(fill="orange", bins=40)+
theme_minimal()
```
## Fuel types and Prices
All fuel types does not cost similar price at any given time as many differnt factors account for the pricing. For example, LPG is one of cheapest fuel while PULP 98 is the expensive fuel type. 

Average pricing of all fuel types.

```{r}
#Fuel prices by fuel type
ggplot(fueldata,
aes(x=Fuel_Type, y=Price,
fill=Fuel_Type))+
geom_boxplot()+
theme_minimal()+
coord_flip() 
```
### Trend of FUEL Price by month

```{r}
# monthly prices 
fueldata <- fueldata |>  # update data with the new column month
  mutate(month = month(date))

monthly <- fueldata |> 
  group_by(month, Fuel_Type, colour=Fuel_Type) |> 
  summarise(
AveragePrice=mean(Price)
)

ggplot(monthly,
aes(month,
AveragePrice,
))+
 geom_point(aes(colour=Fuel_Type))+
    geom_smooth()+
    theme_classic()
```

## Making interactive
```{r}
my_plot <- ggplot(monthly,
aes(month,
AveragePrice,
colour= Fuel_Type,
frame= month))+
#geom_line(color="blue")+
geom_point()+
theme_minimal()

ggplotly(my_plot)
```

