---
title: "Report_ben"
format: html
editor: visual
categories: [R, 2025 Winter]
date: Jul-2025
---

## The goal: To find out the worst book(in English) published in the last 20 years based on Goodreads data set.


# Loading Dataset and library


```{r}
books <- read.csv(file = "data/books.csv")
#| message: false
#| results: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)


```

# Looking at the dataset

```{r}
#| echo: true
ggplot(books, aes(y = language_code)) + 
  geom_bar(fill = "blue") +
  theme_bw()+
  labs(title = "Language of books in the dataset")
```

We have found that the datasets countain books from different languages.

# Filter the books in English and combine them

```{r}
books_eng <- books %>% 
  mutate(Eng = language_code == "eng" |
           language_code == "en-US" |
           language_code == "en-CA" |
           language_code == "en-GB") %>% 
  filter(Eng == 1)

books_eng %>% 
ggplot(aes(y = language_code)) + 
  geom_bar(fill = "tomato") +
  theme_bw()+
  labs(title = "Plot checking filter worked")
```
Based on the website
<https://www.goodreads.com/list/show/24328>. filter rating <3.6, at least >100 ratings count.
```{r}
books_eng_cleaned <- books_eng %>% 
  filter(average_rating < 3.6 , publication_date > 2005, ratings_count >100) 
```

# And the Worst book in the past 20 year is...

```{r}
rating <- books_eng_cleaned %>% 
  filter(average_rating == min(average_rating))

print(paste("The lowest rating book is", rating$title, "by", 
      rating$authors))

people <- books_eng_cleaned %>% 
  filter(ratings_count == max(ratings_count))

print(paste("The book that most people hate is", people$title, "by", people$authors))
```

And to find out which publisher got the most amount of bad books...

```{r}
books_eng_cleaned %>% 
ggplot(aes(x = publisher)) + 
  geom_bar(fill = "cyan") + 
  theme_linedraw()+ 
  labs(title = "Worst publisher")
```

This graph shows that we need to filter the number of publishers

```{r}
publishers <- books_eng_cleaned %>%
  group_by(books_eng_cleaned$publisher) %>% 
  filter(n() > 5)
  
 publishers %>% 
    ggplot(aes(y = publisher)) + 
    geom_bar(fill = "purple") + 
    theme_linedraw()+ 
  labs(title = "Worst publisher filtered")

```

