---
title: "Bird Strike Dashboard"
author: Ediana
date: Jan-2025
format: dashboard
# runtime: shiny # (turned of because no server context used in chunks)
categories: [R, 25Summer, "data: birds_strikes.csv"]
---  

```{r setup, include=FALSE}
# library(shiny)
library(dplyr)
library(tidyr)
library(DT)
library(plotly)

# Load the dataset
birds_strikes <- read.csv("data/birds_strikes.csv")

#| echo: false
```

## Overview

### About the Dataset
This dashboard presents an analysis of **bird strike incidents** reported in the **United States** during the **1990s**, focusing on **various airlines**, **aircraft types**, and **bird species**. The dataset encompasses details on the **impact of bird strikes**, **phases of flight** during which incidents occurred, and **associated geographical** locations.


## Bird Species Involved


### Top 10 Bird Species by Frequency

```{r}
#| echo: false

bird_species <- birds_strikes %>%
  count(species) %>%
  arrange(desc(n)) %>%
  head(10)

plot_ly(bird_species, x = ~species, y = ~n, type = 'bar', name = 'Bird Species') %>%
  layout(title = 'Top 10 Bird Species Involved in Strikes',
         xaxis = list(title = 'Species'),
         yaxis = list(title = 'Frequency'))

```

## Effect of Strikes


### Distribution of Effects

```{r}
#| echo: false
effects <- birds_strikes %>%
  count(effect) %>%
  drop_na()

plot_ly(effects, labels = ~effect, values = ~n, type = 'pie') %>%
  layout(title = 'Distribution of Strike Effects')

#| echo: false
```

## Phases of Flight


### Strikes by Phase of Flight

```{r}
#| echo: false
phases <- birds_strikes %>%
  count(phase_of_flt) %>%
  drop_na()

plot_ly(phases, x = ~phase_of_flt, y = ~n, type = 'bar', name = 'Phase of Flight') %>%
  layout(title = 'Bird Strikes by Phase of Flight',
         xaxis = list(title = 'Phase of Flight'),
         yaxis = list(title = 'Frequency'))

```

## Geographical Distribution


### Strikes by State

```{r}
#| echo: false
states <- birds_strikes %>%
  count(state) %>%
  drop_na() %>%
  arrange(desc(n))

DT::datatable(states, options = list(pageLength = 10))
```
