---
title: "Coffee Project"
author: "Michael"
format: html
editor: visual
warning: false
error: false
date: Jan-2026
categories: [R, 26Summer, "data: coffee_survey.csv"]
---

# Coffee Consumption Group Project

## 1. Install The Libraries:

<!-- Check how to hide install notes and warnings -->

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel) 
library(tidyverse)
library(stringr)
library(plotly)
```

## 2. Install The Data:

```{r}
coffee <- read.csv("../../../../data/coffee_survey.csv")
```

## 3. Explore the data columns:

```{r}
names(coffee)
```

Now we can select some coffee-related columns to explore people's preferences

Let's start with **favourite** type of coffee:

```{r}
unique(coffee$favourite)
```

**Age** is also a useful variable to explore

```{r}
unique(coffee$age)
```

Unfortunately this data is poorly formatted...

## 4. Prepare the data

#### Clean up the **favourites** column

```{r}
coffee <- coffee %>% 
    mutate(favourite = str_replace_all(favourite, c(
      "Blended drink \\(e.g. Frappuccino\\)" = "Blended", 
      "Regular drip coffee" = "Drip")))   
```

#### Clean up the **age** column

```{r}
coffee <- coffee %>% 
    mutate(age = str_replace_all(age, c(
      " years old" = "", 
      ">65" = "65+")))  
```

#### Clean up the **cups** column

```{r}
coffee <- coffee %>% 
    mutate(cups = str_replace_all(cups, c(
      "More than 4" = "5+", 
      "Less than 1" = "<1")))  
```

## 5. Visualise Coffee Preferences

### First let's see the distribution of favourite coffee types:

```{r}
ggplot(data = coffee,
           mapping = aes(y = favourite)) + 
      geom_bar()
```

There's still a lot going on here... We can simplify the data by grouping coffee types into broader preperation styles.

### Group by **Preperation Style**

```{r}
 PrepStyle <- coffee %>% 
      mutate(prep = case_when(
        favourite %in% c("Mocha", "Latte", "Cappuccino", "Cortado", "Espresso")
        ~ "Espresso",
        favourite %in% c("Iced coffee", "Cold brew", "Blended") 
        ~ "Iced",
        favourite %in% c("Drip", "Pourover", "Americano") 
        ~ "Brewed",
        .default = "other"))
```

### Now we can visualise these groups

```{r}
ggplot(data = PrepStyle,
           mapping = aes(x = prep)) + 
      geom_bar()
```

### Reorder Preperation Styles

```{r}
PrepStyle$prep <- factor(PrepStyle$prep, levels=c(
  'other', 'Iced', 'Brewed', 'Espresso')
  )
```

You can arrange the order of attributes to best visualise the data you are working with

### Stacked Bar Graphs

Now we can add **stacks** to the bar graphs to explore relationships between preperation and other attributes

#### Preperation by Age

```{r}
ggplot(data = PrepStyle,
           mapping = aes(x=age, fill=prep)) + 
      geom_bar(position='stack') +
      geom_text(aes(label = after_stat(count)), 
                stat = "count", position = "dodge", check_overlap = TRUE
      )
```

```{r}
ggplot(data = PrepStyle,
      aes(fill=prep, x=age)) + 
  geom_bar(position='dodge') +
  geom_text(aes(label = after_stat(count)),
    stat = "count", position = position_dodge(.9)
  )
```

#### Preperation by Cups

```{r}
ggplot(data = PrepStyle,
           mapping = aes(x=cups, fill=prep)) + 
      geom_bar(position='stack')
```

```{r}
ggplot(data = PrepStyle,
       aes(fill=prep, x=cups)) + 
  geom_bar(position='dodge')
```

#### Preperation by Roast

```{r}
ggplot(data = PrepStyle,
           mapping = aes(x=roast_level, fill=prep)) + 
      geom_bar(position='stack')
```

#### Preperation by Style

```{r}
ggplot(data = PrepStyle,
           mapping = aes(y=style, fill=prep)) + 
      geom_bar(position='stack')
```

<!-- ## 6. Espresso or Brew? -->

<!-- Do younger people drink more espresso coffee? -->

<!-- ```{r} -->

<!-- PrepStyle <- PrepStyle |>  -->

<!--       mutate(AgeGroup = case_when( -->

<!--         age %in% c("<18", "25-34", "18-24") ~ "Young", -->

<!--         .default = "old")) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- PrepStyle <- PrepStyle |>  -->

<!--       mutate(AgeGroup = age == "<18", "25-34", "18-24") -->

<!-- ``` -->

<!-- ```{r} -->

<!-- t.test(prep ~ AgeGroup, data = PrepStyle) -->

<!-- ``` -->
