---
title: "Melbourne Housing"
date: today
format: dashboard
warning: false
---
##
<div style="text-align:center;">
Objective: Identify factors affect selling price of houses  
Details of Data set ::  
No of factors              :  20  
No of Quantitative factors :  10  
No of Qualitative factors  :	 9  
No of records	             :  13580  
</div>

## {height=40%, width=50%}

```{r}
#| message: false
#| warning: false
#| width: 30%
library(knitr)
table1 <- read.csv("Tables/tables1_1.csv")
kable(table1)
```

```{r}
#| message: false
#| warning: false
#| width: 70%
# library(magick)
table2 <- read.csv("Tables/table2.csv")
kable(table2)
```

## 

```{r}
#| message: false
#| warning: false

library(patchwork)
library(ggplot2)
library(plotly)

housing <- read.csv("Tables/melb_data.csv")

housing$Suburb<-as.factor(housing$Suburb)
housing$Rooms_1<-as.factor(housing$Rooms)
housing$Type<-as.factor(housing$Type)
housing$Method<-as.factor(housing$Method)
housing$Postcode<-as.factor(housing$Postcode)
housing$Bedroom_2<-as.factor(housing$Bedroom2)
housing$Bathroom_2<-as.factor(housing$Bathroom)
housing$CouncilArea<-as.factor(housing$CouncilArea)
housing$Regionname<-as.factor(housing$Regionname)
housing$Propertycount<-as.factor(housing$Propertycount)
housing$Car_2<-as.factor(housing$Car)

p1<-ggplot(housing ) + 
  geom_boxplot(aes(x=Bedroom_2,y=Price, fill=Bedroom_2))+
  theme_classic()


p2<-ggplot(housing ) + 
  geom_boxplot(aes(x=Bathroom_2,y=Price, color=Bathroom_2))+
  theme_classic()

p3<-ggplot(housing ) + 
  geom_boxplot(aes(x=Rooms_1,y=Price, color=Rooms_1))+
  theme_classic()


p4<-ggplot(housing ) + 
  geom_boxplot(aes(x=Car_2,y=Price, color=Car_2))+
  theme_classic() 


p5<-ggplot(housing ) + 
  geom_boxplot(aes(x=Type,y=Price, color=Type))+
  theme_classic()


p6<-ggplot(housing ) + 
  geom_boxplot(aes(x=Method,y=Price, color=Method)) +
  theme_classic()

p1+p2+p3+p4+p5+p6+ plot_layout(ncol = 2)
```

```{r}
#| message: false
#| warning: false
#| echo: false
library(dplyr)
housing_2<-housing |>
  select(Price,Rooms, Distance,Bedroom2,Bathroom,Car, Landsize, BuildingArea)
library(PerformanceAnalytics)
housing_3<- na.omit(housing_2)
plt1ly1 <- chart.Correlation(cor(housing_3))
ggplotly(plt1ly1)
```

## {width=30%}

```{r}
#| message: false
#| warning: false
#| width: 30%
# library(magick)
table3 <- read.csv("Tables/table3.csv")
kable(table3)
```

##
Issues:  
1. Two columns are number of rooms.  
2. Some houses are with 0 rooms.  
3. 20 rooms are in a house.  
4. More columns contain NAs.  
5. More quantitative variables(factors) are inter-correlated.  
6. Place related factors/variables are connected/compounding.           