---
title: "Cafe"
author: "Amber Sewell-Green"
# format: pdf
editor: visual
date: 28-Jan-2026
categories: [R, 26Summer, "data: coffee_survey.csv"]
---

\<!--The `echo: false` option disables the printing of code (only output displayed). \<!-- CNTR + ALT + I to create new code chunks

## PROJECT CAFE

I'm using the ../../../../data/coffee_survey.csv and below is my analysis. Let's take a closer look!

This is a survey of 3790 consumers who were asked 43 questions about their coffee preferences

<!--# This CNT + SHIFT + C HIDES TEXT -->

```{r}
#| echo: false
cafe <- read.csv("../../../../data/coffee_survey.csv")
```

```{r}
#| echo: false
head(cafe)
names(cafe)
str(cafe)
summary(cafe)
```

```{r}
#| echo: false
library(dplyr)
library(tidyr)
nrow(cafe)
```

```{r}
#| echo: false
ncol(cafe)
```

## CLEANING THE CUPS (Data Cleaning)

Let's look at a few interesting factors and remove missing values

```{r}

#Removing the NA's from the variables I am interested in as a new df called "clean_cup"
unique(cafe$cups)
library(tidyr)
clean_cup <- cafe |> 
  drop_na(political_affiliation, number_children, cups, favourite, strength, caffeine, expertise, total_spend, education_level, employment_status)

#other way to remove NA
#clean_cup <- cafe |>
 # filter(!is.na(cups),
        # !is.na(employment_status))


```

##BULDING THE CAFE (Changing the Data)

Let's continue to make the data in a format we can use by changing the column names, replacing characters and making characters/factors into numerical data

```{r}

#First change the names of the columns to make the code easier to use: politics, kids, cups, caffeine

clean_cup <- clean_cup %>% 
  rename(politics = political_affiliation)

clean_cup <- clean_cup %>% 
  rename(kids = number_children)

#clean_cup$politics
#unique(clean_cup$politics)
#summary(clean_cup$politics)

#clean_cup$kids
#unique(clean_cup$kids)
#summary(clean_cup$kids)

#clean_cup$caffeine
#unique(clean_cup$caffeine)
summary(clean_cup$caffeine)

#clean_cup$cups
#unique(clean_cup$cups)
summary(clean_cup$cups)

library(ggplot2)
install.packages("stringr")
library(stringr) #use to change characters quickly
library(tidyverse)

#Next let's change the order of "number of cups consumed" to be ordinal/incremental


#Make a smaller more manageable data frame called small_cafe

small_cafe <- clean_cup |> 
  select("politics", "cups", "caffeine", "kids")

#For number kids and cups make categorical variables numerical

#Change to numbers: if its this turn to that if else leave as is
small_cafe <- small_cafe |> 
  mutate(kids = if_else(kids == "None",
                        true = "0",
                        false = kids),
         kids = if_else(kids == "More than 3",
                        true = "4",
                        false = kids))

#Another way to do the same thing by make it a case when this = this do this.
 
small_cafe <- small_cafe |> 
  mutate(kids = as.numeric(case_when(kids == "None" ~ "0", 
                          kids == "More than 3" ~ "4",
                          .default = kids)))

#Apply to the "cups" data

small_cafe <- small_cafe |> 
  mutate(cups = if_else(cups == "Less than 1",
                        true = "0",
                        false = cups),
         cups = if_else(cups == "More than 4",
                        true = "5",
                        false = cups))
#Other Option
 
small_cafe <- small_cafe |> 
  mutate(cups = as.numeric(case_when(cups == "Less than 1" ~ "0", 
                          cups == "More than 4" ~ "5",
                          .default = cups)))
#Make data numerical

small_cafe <- small_cafe |> 
  mutate(kids = as.numeric(kids),
         cups = as.numeric(cups))

# if(!is.numeric(small_cafe$cups)) small_cafe$cups <- as.numeric(small_cafe$cups)

class(small_cafe$cups) #santity check the class is numeric
class(small_cafe$kids)
small_cafe$kids <- as.numeric(small_cafe$kids)#same
```

```{r}

#Make cups & kids data ordinal
#small_cafe2 <- small_cafe$cups <- factor(small_cafe$cups, levels = c(0, 1, 2,3, 4, 5)) #!!! ADDS NA BACK IN

#view(small_cafe2)
```

## DESCRIBE THE ROAST (Descriptive Statistics)

We can see most people in this data set drink 1-5+ cups of coffee daily, the average being 2 cups a day.

Meanwhile most people have 0-4+ kids, with the majority having no kids.

```{r}
# Descriptive Statistics
median(small_cafe$cups)
min(small_cafe$cups)
max(small_cafe$cups)
range(small_cafe$cups)
diff(range(small_cafe$cups))

median(small_cafe$kids)
min(small_cafe$kids)
max(small_cafe$kids)
range(small_cafe$kids)

     
#Quartiles
quantile(small_cafe$cups)
IQR(small_cafe$cups)
quantile(small_cafe$kids)
IQR(small_cafe$kids)

#SD & Variance
sd(small_cafe$cups)
var(small_cafe$cups)
sd(small_cafe$kids)
var(small_cafe$kids)

#Statistical Summary
summary(small_cafe$cups)
summary(small_cafe$kids)

#2. Measures of Correlation - more than one set of data----

#Covariance
cov(small_cafe$cups, small_cafe$kids)

#Correlation

#Pearson
cor(small_cafe$cups, small_cafe$kids)

```

## SEE THE AROMA (PLOTTING)

Let's take a closer look at how political preferences and family number influence caffeine intake.

\<!--\> Does family size (kids) influence caffeine intake?

\<!--\> Does political preference (politics) influence caffeine intake?

```{r}
# Visualise in ggplot
library(RColorBrewer)


ggplot(data = small_cafe)+ # data = your data's name
  (mapping = aes(x = cups, #x variable
                 y = kids)) + #y variable
  geom_bar(stat = "identity",
           aes(fill = as.factor(cups))) + 
  #scale_fill_brewer(palette = "Accent") +#plot type
  theme_classic() #theme

```