---
title: "Wildlife"
author: "Nayomi and Julian"
format: html
editor: visual
date: Jan-2026
categories: [R, 26Summer]
---

## Introduction

The database we used here contains records of reported wildlife strikes since 1990. this contains self reported strikes from airlines, airports, pilots, and other sources.

## Import the data set and install data packages

```{r}
wildlife_raw <- read.csv('data/wildlife_impacts.csv')
library(dplyr)
library(ggplot2)
library(knitr)
```

This is how the data set looks like

```{r}
head(wildlife_raw)
Table1 <- head(wildlife_raw)
kable(Table1)
```

Let's see the trend over years Is this really increasing over time or proportionately the same???

```{r}
ggplot(wildlife_raw,
       aes(x=incident_year,))+geom_bar()+labs(x="Year of the incident", y="Number of Strikes")
```

## Data purification

There were a lot of NAs in the data set and treated accordingly.

e.g.

```{r}
wildlife_cor1 <- wildlife_raw %>% 
  filter(!is.na(speed))
wildlife_cor1 <- wildlife_cor1 %>% 
  filter(!is.na(damage))
```

## Analysis

## Differences in strike count according to month of year

```{r}
ggplot(wildlife_raw,
       aes(x=incident_month, colour=operator))+geom_bar()+
  labs(x="Month", y="Number of Strikes")
```

## Histogram number of strikes at height

```{r}
wildlife_impacts <- read.csv("data/wildlife_impacts.csv")
wildlife_impacts <- wildlife_impacts %>%
  filter(!is.na(height))
wildlife_impacts <- wildlife_impacts %>%
  filter(!is.na(time_of_day))
ggplot(data = wildlife_impacts,
       aes(x = time_of_day, y = height)) + 
  geom_boxplot(colour = "black", fill = "gray") + 
  labs(x = "time of day", y = "height")
```

## Time of day and strike count

```{r}
library(ggplot2)
wildlife_impacts <- wildlife_impacts %>%
  filter(!is.na(time_of_day))
ggplot(data = wildlife_impacts,
       aes(x = time_of_day)) + 
         geom_bar(fill = "gray", colour = "black") + 
  labs(x = "time of day", y = "number of strikes")
```

## Histogram of time of day and height of strike

```{r}
ggplot(data = wildlife_impacts,
       mapping = aes(x = height)) + 
  geom_histogram(bins = 24, colour = "black", fill = "gray")+ 
  labs(y = "number of strikes")
```

## Stats

```{r}
summary(wildlife_impacts$height)
model <- lm(formula = "speed ~ cost_repairs_infl_adj", data = wildlife_impacts)
summary(model)
wildlife_impacts <- wildlife_impacts %>%
  filter(!is.na(speed))
wildlife_impacts <- wildlife_impacts %>%
  filter(!is.na(cost_repairs_infl_adj))
library(scales)
ggplot(wildlife_impacts, aes(x = speed, y = cost_repairs_infl_adj)) +
  geom_jitter() +
  scale_y_continuous(labels = dollar) +
  labs(y = "cost of repairs")

```
