---
title: "Good Reads"
author: "Aisha Aslam"
format: html
code-fold: true
code-summary: "Show code"
code-tools: true
#editor: visual
date: Jul-2025
warning: false
categories: [R, 25Winter, "data: books.csv"]
---

## Introduction

If you love books, you are at just the right place :) Wouldn't it be interesting to find out:

***Is there a correlation between higher book ratings and the number of book reviews some popular books receive?***

Let us find out using the following dataset.

## The Dataset

We will be using the publicly accessible dataset 'Good reads books' . (<https://www.kaggle.com/datasets/jealousleopard/goodreadsbooks> ).

```{r}
#| eval: false
#| include: false
install.packages(c("dplyr", "ggplot2", "plotly"))
install.packages("knitr")
```

```{r}
library("dplyr")
library("ggplot2")
library("plotly")
library("knitr")
```

**A Sneak-peak into the dataset**

```{r}
goodreadsdata <- read.csv("../../../../data/books.csv")

kable(head(goodreadsdata))
```

**Let's visualize the book reviews and book ratings**

```{r}
#View(goodreadsdata)

ggplot(goodreadsdata, aes(x = average_rating, y = text_reviews_count)) +
  geom_point( colour = "blue") 


```

More ratings = more reviews :)