---
title: "Book Report"
author: "Liz"
format: html
editor: visual
date: Jul-2025
categories: [R, 2025 Winter]
---

## Data set

We took a deep dive into our books data set. We have captured authors, publishers, and ratings.

## Exlporing the data set

First we told R to use dplyr
```{r}
#| label: Loading Library
#| message: false
library(dplyr)
```

Then we told R to look at our dataset
```{r}
#| eval: false
read.csv("data/books.csv")
```

Then we asked R to make an object called Books

```{r} 
Books <- read.csv("data/books.csv")
```

Next we asked a series of questions about the data:

1.  What is the range of page numbers?

```{r}
range(Books$num_pages)
```

2.  What is the range of rating counts?
```{r}
#| message: false
#| warning: false
range(Books$ratings_count)
```

3.  What publishers were used in the data set?

```{r}
library(ggplot2)
```

```{r}
??Books
```

```{r}
#| message: false
#| warning: false
str(Books)
```

First, using a scatterplot:
```{r}
ggplot(data = Books,
       mapping = aes(x = X,
                     y = publisher))+
  geom_point()
```

Second, using a bar graph:
```{r}
#| echo: true
ggplot(data=Books,
       mapping=aes(x=publisher))+
  geom_bar()
```

Third, we were only interested in publishers of books with less than 5 pages

```{r}
Books%>%
    filter(num_pages<5)%>% 
    ggplot(mapping=aes(y=publisher))+
  geom_bar()
```

Fourth, we were interested in books published by Scholastic

```{r}
Books%>%
    filter(grepl("Scholastic", publisher))%>% 
    ggplot(mapping=aes(y=publisher))+
  geom_bar()
```
4.  Are rating counts associated with number of pages?

```{r}
cor.test(Books$num_pages,Books$ratings_count)
```
5.  Compare ratings counts for books in English versus books not in English?

First we must create a categorical variable English vs. Non-English

```{r}
Books <- Books |>
  mutate(english_books=language_code=="eng")
```

Now we can run a t-test to compare ratings counts

```{r}
t.test(ratings_count~english_books,data=Books)
```
Now lets visualise it

```{r}
library(ggplot2)
```

```{r}
ggplot(Books,
       aes(x=english_books,y=ratings_count,
           fill=english_books))+
  geom_col()
```

