---
title: "Workshop Project Report"
author: "Andrew"
format: html
date: Jul-2025
categories: [R, 25Winter, "data: melb_data.csv"]
---

## Analysing a dataset in R

We are using the **png**, **ggpubr**, **tidyverse** and **plotly** libraries to examine our data. We can install and enable these libraries as follows, using an if loop to prevent repeat installation. 

```{r}
#| message: false
inst_list = c("tidyverse", "plotly", "png", "ggpubr")

for(i in inst_list){
  if(!inst_list[i] %in% installed.packages()){
    print(inst_list[i])
    install.packages(inst_list[i])
  }
}

library(png)
library(ggpubr)
library(tidyverse)
library(plotly)
```


## Melbourne Housing Data

The dataset we have chosen is the Melbourne Housing Dataset. We can import the data and run a summary as follows: 

```{r}
melb_data_raw <- read.csv("data/melb_data.csv")
summary(melb_data_raw)
```

Data has been imported to a 'raw' data object, to be drawn from to produce usable data. 

## Data Cleaning

This data includes some values we would like to change before we continue, so we can load the data into a new object for manipulation.
From the summary, we can see that the oldest house was built in 1196. Since Melbourne was settled in 1835, this datapoint is a clear outlier and suggests it may be a typo. Therefore we can mutate this datapoint as we load the data into a new object:

```{r}
melb_data <- melb_data_raw %>% mutate(YearBuilt = 
                                        ifelse(YearBuilt < 1800,NA,YearBuilt))
```

No other clear outliers/typos exist. Landsize of 0 appears to relate to apartments. Postcode, latitude, longitude, distance all within reasonable bounds. 

## Exploring Data

To explore the data, we can create an object called plot_map to store a ggplot of the data, using the latitude and longitude along the x and y axis. This can then be called with geom_point() to produce a plot. 

```{r}
plot_map <- ggplot(data = melb_data, mapping = aes(x = Longtitude, y = Lattitude))

plot_map + geom_point(mapping = aes(colour = YearBuilt)) + 
              theme_classic() + scale_color_viridis_c()
```

This graph uses the latitude and longitude attributes of the dataset to produce a scatterplot of all house sales in Melbourne, the sum of these data points approximates the geography of Melbourne. The colours can show some hotspots for builds during certain years. 

We can overlay this graph on a map of melbourne to show how the areas relate to the real world by taking a map of Melbourne from google and using it as a background image for the graph. This is read in using the png library, and limits are set on the x,y coords of graph to fit image:

```{r}
map_img <- png::readPNG("gallery/R/25Winter/Andrew Proj/data/map_desaturated.png")

plot_map + background_image(map_img) + geom_point((mapping = aes(colour = 
              YearBuilt))) + theme_classic() + scale_color_viridis_c() + 
              coord_cartesian(xlim = c(144.4,145.7), ylim = c(-38.2, -37.4))
```

This can alternatively be done using ggmap() rather than an image for the background, however this requires API access. 

This data can be aggregated by region as follows:

```{r}
tooltip_data <- melb_data %>% 
  group_by(Regionname) %>% 
  summarise(Latitude = median(Lattitude), Longitude = median(Longtitude), Houses = 
              sum(Type == "h"), Townhouses = sum(Type == "t"), Units = 
              sum(Type == "u"), Properties = n(), Mean_Price = median(Price), PropertySize = median(Landsize))
```

```{r}
#| echo: false
tooltip_data
```

This table separates out the median price, latitude, longitude, number of houses/units/townhouses and land size of properties. 

The goal was then to use these in plotly to have hoverable aggregated plot points, however I wasn't able to finish this.

```{r}
tooltip_map <- ggplot(data = tooltip_data, mapping = aes(x = Longitude, y = Latitude)) + background_image(map_img) + geom_point(data = tooltip_data, label = tooltip_data$Regionname, label2 = tooltip_data$Mean_Price, label3 = tooltip_data$Houses, label4 = tooltip_data$Units) + theme_classic() + scale_color_viridis_c() + 
  coord_cartesian(xlim = c(144.4,145.7), ylim = c(-38.2, -37.4))

tooltip_map
ggplotly(tooltip_map)
```

```{r}
melb <- read.csv("data/melb_data.csv")
melb |>
  summary() |>
  knitr::kable()
```

