---
title: Analysing your own data
--- 

In this final session, we look at taking the skills we've developed during the intensive and applying it to your own data. We'll also discuss a few common pitfalls and where to get help in the wild.

## Setting up

Spend five minutes locating your data and preparing Spyder

1. Locate and structure your own project folder
    * You might like to create a *data/* folder
    * You might like to create a *scripts/* folder
2. (Optional) Turn it into an RStudio project
    1. Click *File* $\rightarrow$ *New Project...*
    2. Click **Existing Directory**
    3. Click *Browse...* and choose the ***top-level folder*** of your project
3. Create a new script for processing, called `processing.R` (or whatever you'd like to).

We'll spend most of this session like project time, troubleshooting and setting up your own data for analysis. Before we do that, let's discuss a few common tips and pitfalls. Namely,

* Managing paths and the working directory
* Importing different data structures and files
* Dealing with different types of data

## Environments and importing data

Perhaps the most frustrating issues are those which prevent you from importing the data at all!

### Getting your setup right
Nobody likes to encounter this:

```{r}
#| echo: false
#| error: true
read.csv("PANIC!")
```

To solve it, we need to talk about filesystems.

When running an R script, there are always three relevant locations:

* Your R executable
* Your script
* Your working directory

The R executable runs your script **from the working directory**

Why does this matter? Because when you import data, R uses your **working directory** as the reference, not the others. This means that there are important questions you need to ask:

1. Where is your working directory?
2. Where is your data?

Answering the first question is easy: simply run

```{r}
getwd()
```

:::{.callout-note}
## Alternatively

* When you run R from the command line, it's the current location as specified in the terminal
* When you run R in RStudio, it's the address displayed in the bar attached to **Console**
:::

This address is your working directory. All paths in your scripts are evaluated relative to this location.

This includes data paths, which can be absolute or relative, and can be online.

Absolute paths begin with a root prefix and have folders separated with slashes. They contain **all** the folders from the root of the filesystem down to the object of interest.

:::{.panel-tabset group="os"}

## Windows

On Windows, absolute paths are prefixed with the drive, e.g. `C:\` and folders separated with backslashes `\`

```
C:\Users\...\...\data\example.csv
```

## Unix-like (e.g. macOS, Ubuntu)

On Unix-like systems, absolute paths are prefixed with a forwardslash `/`, which also separate folders.

```
/home/user/.../.../data/example.csv
```

Alternatively, you can start from your 'user' directory by prefixing with a tilde `~`:

```
~/.../.../data/example.csv
```

## URL

Website and web-hosted files can typically be accessed with URLs. Full or 'absolute' URLs are prefixed with a protocol (e.g. `https://`) and a hostname (e.g. `www.website.com`), with folders then separated by forwardslashes

```
https://www.website.com/.../.../.../data/example.csv
```

:::

Relative filepaths have no prefix. 

:::{.panel-tabset group="os"}

## Windows

On Windows, relative paths are still separated with backslashes

```
data\example.csv
```

## Unix-like (e.g. macOS, Ubuntu)

On unix-like systems, relative paths are still separated with forwardslashes

```
data/example.csv
```

## URL

It's possible to have a relative path to a web file, however, as a relative filepath, you **must be running Python from the server.**

The syntax is the same as unix-like systems, i.e., folders separated with forwardslashes

```
data/example.csv
```

:::

:::{.callout-note}
# Should I use absolute or relative paths?

+----------+---------------------------------------------------------+--------------------------------------------------------------------+
|          | Pros                                                    | Cons                                                               |
+----------+---------------------------------------------------------+--------------------------------------------------------------------+
| Absolute | * Works for any working directory (on same device)      | * Only valid on one device                                         |
|          |                                                         | * Can get long                                                     |
|          |                                                         | * Can contain irrelevant information                               |
+----------+---------------------------------------------------------+--------------------------------------------------------------------+
| Relative | * Works on any device with the same *project* structure | * Working directory must be set up correctly                       |
|          | * Only contains project-specific information            | * Can become confusing with many parent folders (e.g. `../../../`) |
|          | * Can be shorter                                        |                                                                    |
+----------+---------------------------------------------------------+--------------------------------------------------------------------+
:::

Once you have your working directory and your filepath, you can now check that any data paths have been specified correctly.

**If the path is absolute**

You just need to ensure that your working directory is on the same device as the file. 

**If the path is relative**

Go to the working directory and trace the path to ensure it's correct. The path begins *inside* the working directory. A few oddities:

* `..` indicates go *up* a folder
* `.` indicates the current folder

:::{.callout-note}
# Changing the working directory

If you're using a project, you shouldn't need to change working directories. Instead, try modifying the path first.

If you really need to change working directories, you can do this with R code,

```r
setwd("path/to/dir")
```

Or in the **Files** tab by navigating to the folder and clicking *More* $\rightarrow$ *Set As Working Directory*.

You should check that it's worked by running `getwd()`.
:::

:::{.callout-warning}
# Watch out Windows users

The backslashes in Windows paths conflict with R's escape character in strings (also a backslash). To fix this, you can

* Replace backslashes with forwardslashes **in R**
* Use *raw strings*: `r"(...)"`
* Escape the backslashes with an *extra* backslash

For example, the following Windows path

<i>
```bash
C:\Users\me\data\secrets.csv
```
</i>

could be imported as any of the following

```r
# Raw strings
read.csv(r"(C:\Users\me\data\secrets.csv)")

# Replace with forwardslashes
read.csv("C:/Users/me/data/secrets.csv")

# Escape the backslashes
read.csv("C:\\Users\\me\\data\\secrets.csv")
```

We recommend using the `r"(...)"` option where possible, as it's the least work.
:::

If you've fixed the filepath and you're on top of the Windows peculiarities, then check the following errors for more troubleshooting.


:::{.callout-warning collapse="true"}
# `Error: ... is an unrecognized escape in character string

You've probably used a path with backslashes and not adjusted it for R. See "Watch out Windows users" above.

:::

:::{.callout-warning collapse="true"}
# `Error: '\U' used without hex digits in character string...`

You've probably used a path with backslashes and not adjusted it for R. See "Watch out Windows users" above.

R throws this error when it runs `"\u..."` or `"\U..."` (unless `"..."` is a valid unicode code; `\Users` is not).

:::

:::{.callout-warning collapse="true"}
#   `cannot open file... ` but everything is correct???

Have you used a path with backslashes on a non-Windows machine? If you have, replace them with forwardslashes.
:::

Final thoughts

* Avoid spaces
* Use relative filepaths where possible
* Get familiar with your working directory

### Importing your data correctly

Once you've got the path working, the next challenge will be importing the data correctly. Unlike our data, yours might have multiple header rows, missing data, simply be organised differently or even be a different file type.

We'll look at importing .csv files here, but the same applies to other file types.

The documentation for `read.csv` describes (as of v4.4.0) 25 different parameters that the functions supports. These include

| Parameter | Description | 
| --- | --- |
| `file` | The path (**doesn't have to actually be `.csv`!**) | 
| `header` | `TRUE/FALSE` specifying if the first row is a header | 
| `sep` | The separator (use `"\t"` for tab separated values) |
| `na.strings` | A list of values to interpret as empty cells | 
| `skip` | Skip the first `skip=n` rows | 

There are plenty more! Run

```r
help(read.csv)
```

to see the rest. For more advanced reading, you can also try the [`readr` package](https://readr.tidyverse.org/).

:::{.callout-tip}
# MS Excel files (`.xlsx`)

Reading `.xlsx` files can be complicated. You should use the [`readxl` package](https://readxl.tidyverse.org/)

```r
library(readxl)
read_excel(...)
```
You can use the argument `sheet="..."` if you'd like to specify the sheet.

[Check out the documentation](https://readxl.tidyverse.org/reference/read_excel.html) for details.
:::


## Dealing with different and dodgy data

Our data has been set up to be a bit of a challenge, and a bit of a help. Your data might be organised differently, and might need more work! You might also need to perform different tasks.

We'll look at a few common tips to get you going, but before you start, the best advice is to get out a pencil and paper and draw. Mock up your data, figure out what you want and write down the steps that you would have to do *by hand*. Then you'll have a good grasp of what you want, and whether the code is working.

A few resources that you should consult:

* [Our compilation of useful resources](https://uqlibrary.github.io/technology-training/R/usefullinks.html)
* [R and RStudio cheatsheets](https://posit.co/resources/cheatsheets/) are fantastic resources which outline the common data tidying tasks along with diagrams of the tabular operations. **Consult these often!**
* [The dplyr documentation](https://dplyr.tidyverse.org/) provides in-depth descriptions and guides for working with dplyr.

See below for a summary of data cleaning tips you might need to apply to your data, assuming your data is stored in `df` and `"col_name"` is a column name

### Reshaping your data

For simple (or complex) reshaping tasks, like filtering, subsetting and adding new columns, refresh yourself with our second session of this intensive: [2 - Data Processing](./Essentials/2-Data_processing.qmd).

The Tidyverse usually wants you to organise your data in the "[tidy data](https://www.jstatsoft.org/index.php/jss/article/view/v059i10/v59i10.pdf)" way: single variable per column, single case per row, one value per cell. To get to that point, we often have to reshape our dataframes, usually from "wide format" to "long format". The package [tidyr](https://tidyr.tidyverse.org/) is perfect for that, using the `pivot_longer()` function (and its `pivot_wider()` opposite).

This tidyr example shows the process in a nutshell:

```{r}
library(tidyr)
# This relig_income dataset has data split into income bracket columns:
relig_income
# To make it more "tidy":
# Store income brackets in a column, so functions can access the data.
reshaped <- relig_income |>
  pivot_longer(!religion, # all columns except "religion"
               # name the columns where data is stored:
               names_to = "income", # what comes from the column headings
               values_to = "count") # what comes from the cells
```

The [tidyr cheatsheet](https://github.com/rstudio/cheatsheets/raw/main/tidyr.pdf) is great to visualise what exactly the `pivot_*()` function do, with helpful pictograms.


<!-- 
* [10 minutes to pandas](https://pandas.pydata.org/pandas-docs/stable/user_guide/10min.html)
* [Indexing and selecting data](https://pandas.pydata.org/pandas-docs/stable/user_guide/indexing.html)
* [Merge, join, concatenate and compare](https://pandas.pydata.org/pandas-docs/stable/user_guide/merging.html)
* [Reshaping and pivot tables](https://pandas.pydata.org/pandas-docs/stable/user_guide/reshaping.html)
 -->

### Cleaning up inconsistencies

The `mutate(...)` function is the simplest for cleaning up values in certain columns. Use as

```r
df <- df %>%
  mutate(new_col = ...)
```

This creates a new column based on the computation of `...`, which can include other columns. A useful helper function is `if_else()`, which changes the computation based on the condition. For example, to replace values `"A"` with `"B"` in `col_name`,

```r
df <- %>%
  mutate(col_name = if_else(col_name == "A", "B", col_name))
```

For reshaping and cleaning, consult the [dplyr cheatsheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf).

### Dealing with different types of data

Each variable (column) in an R dataframe is stored as a particular data type. Common types include

* `character` for literal strings
* `numeric` for numbers
* `int` for integers (no decimals)
* `factor` for categorical data
* `logical` for booleans (`TRUE` or `FALSE`)
* `POSIXct` (among others) for timestamps

For applying methods specific to **textual**, **temporal** or **categorical** data, you must first ensure the columns match the dtype you expect. Common mishaps include

* Timestamps or categorical data stored as `character` types
* Numeric *categorical* data stored as `numeric` the type.

If you need to change the way a variable is stored, you can either do that during the import stage (e.g. by modifying the function `read.csv`) or with a relevant function, described in the respective sections below.

#### Textual

The simplest way to deal with textual data in R is by using the [stringr](https://stringr.tidyverse.org/) package.

:::{.callout-tip}
# Consult the user guide

You should consult the official *[Introduction to stringr](https://stringr.tidyverse.org/articles/stringr.html)* user guide for details about the package, and the [official cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf) for a detailed summary. We've included a brief summary of some useful functions here.
:::

<!-- 
For each example below, append `.method` to `df["col_name"]`. For example, the method `.str.replace` should be used as `df["col_name"].str.replace()`. Each function is applied to **every row in "col_name"**
 -->

Load in the package with `library(stringr)`. All stringr functions follow the same format:

<i>
```r
str_<function_name>(input_string, parameters...)
```
</i>

The value for `input_string` can be a variable(s). If you're using pipes, you could use `select()` to choose the columns to include. For example

```r
df %>%
  select(a_string_col) %>% 
  str_length()   # Or any other stringr function
```

Some useful functions are included here for reference.

| Function | Description |
| --- | --- |
| `str_detect(...)` | Return TRUE/FALSE if pattern is in the string. | 
| `str_sub(...)` | Extract a substring based on character indices (a slice). | 
| `str_trim(...)` | Trim whitespace from start and/or end of the string. | 
| `str_replace(...)` | Replace the first matched pattern in the string (use `str_replace_all(...)` to replace all). | 
| `str_c(...)` | Join multiple strings into one. |


#### Time series
The simplest way to deal with temporal data in R is by using the [lubridate](https://lubridate.tidyverse.org/) package.

:::{.callout-tip}
# Consult the user guide

You should consult the official *[Do more with dates and times in R](https://lubridate.tidyverse.org/articles/lubridate.html)* user guide for details about the package, and the [official cheatsheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) for a detailed summary. We've included a brief summary of some useful functions here.
:::

Load in the package with `library(lubridate)`. To convert the column `dt_col` into an appropriate type, use

```r
df$dt_col <- as_datetime(df$dt_col)
```

There are other parsing functions, should your needs be more specific. [Take a look at the cheatsheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) to learn more.

Most lubridate functions will take in a date-time object (or a variable containing them). You can use the pipe, for example

```r
df %>%
  select(a_datetime_col) %>%
  year()
```

Some useful functions are included here for reference.

| Function | Description |
| --- | --- |
| `hour()` | Returns the hour part of the datetime. Similarly for other units (e.g. `year()`, `second()`). | 
| `hours(...)` | Represents a number of hours (similarly for other units). Useful for arithmetic, e.g. `df$dt_col + hours(2)`. ***Can result in impossible dates***, e.g. 31st Feb.
| `dhours(...)` | Represents a number of hours in terms of seconds (similarly for other units). Useful for arithmetic, e.g. `df$dt_col + dhours(2)`. ***Inconsistent units (e.g. months, leap-years) are approximated***.
| `round_date(...)` | Rounds to the near unit specified. | 
| `with_tz(...)` | Convert a date-time to a new time zone | 

##### Filtering

You can filter particular parts of the timestamp as normal, because they're just numbers:

```r
df %>%
  filter(year(dt_col) == 2026)
```

However, you might want to filter from or before a particular date-time. Use the parsing functions to do this:

```r
df %>%
  filter(year(dt_col) > as_datetime("2026/02/03"))
```

#### Categorical

R has a special, built-in type for categorical data: `factor`. It stores the different possible values as "levels", which it maps to your data, making it more efficient than normal strings.

Factors can also order the levels, making it a good choice for ordered data (e.g. "Small" < "Medium" < "Large").

To make the variable `cat_col` a factor, use the `factor()` function:

```r
df$cat_col <- factor(df$cat_col)
```

You can use `levels(df$cat_col)` to see the unique levels.

For more advanced functions, you can use the [forcats](https://forcats.tidyverse.org/) package. 


:::{.callout-tip}
# Consult the user guide

You should consult the official *[Introduction to forcats](https://forcats.tidyverse.org/articles/forcats.html)* user guide for details about the package, and the [official cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/factors.pdf) for a detailed summary. We've included a brief summary of some useful functions here.
:::

Load in the package with `library(forcats)`. 

All forcats functions follow the same format:

<i>
```r
fct_<function_name>(input_factor, parameters...)
```
</i>

If you're using pipes, you could use `select()` to choose the columns to include. For example

```r
df %>%
  select(a_cat_col) %>% 
  fct_count()   # Or any other forcats function
```

Some useful functions are included here for reference.

| Function | Description |
| --- | --- |
| `fct_count()` | Count the number of values per level |
| `fct_c()` | Combine factors with different levels | 
| `fct_relevel(...)` | Manually reorder the factor levels | 
| `fct_recode(...)` | Manually change the levels | 
| `fct_collapse(...)` | Collapse multiple levels into a manual groups | 
| `fct_other(...)` | Collapse selected levels into a single "other" level | 

#### Geospatial

Managing geospatial data is a complex task beyond the scope of this introductory intensive. If you're planning to analyse geospatial data in R, you should check out

* [sf](https://r-spatial.github.io/sf/articles/sf1.html) (simple features) for **vector data**
* [terra](https://rspatial.github.io/terra/index.html) for **raster data** and more complex vector data
* [stars](https://r-spatial.github.io/stars/) for spatiotemporal data, irregular grids or other advanced features

Let us know if you'd like a hand or want to discuss further!

## What to do when you don't know what to do

1. **Consult the documentation**. If your function isn't behaving, go to the specific page for that function. Otherwise, consult a relevant user guide or [cheat sheet](https://posit.co/resources/cheatsheets/).
2. **Search the web for your issue** to find practical and canonical approaches to typical data cleaning tasks. [StackOverflow is your friend](https://stackoverflow.com/questions/tagged/r).
3. **Ask us while we're here!** And once we're gone, shoot us an email at [training@library.uq.edu.au](mailto:training@library.uq.edu.au)
4. Lots of people also ask AI. There are pros and cons, but beware: you'll get a solution that *probably* works, but if you don't know why, **you should double check the data matches what you want** (and try to understand what it's done!).

## Now give it a go!

For the rest of this session, we'd love to help you get your data set up and working on your end. For that reason, we've dedicated this time to troubleshooting any issues that arise together. Good luck!